---
title: Supplementary Information I. Statistical Analysis of "Bush-crickets show lifelong flexibility in courtship signals to match predation threat"
author: Antoniya Hubancheva (antonia.hubancheva@gmail.com), Viktor Senderov, Manuela Nowotny, Stefan
  Schöneich and Holger R. Goerlitz
output:
  html_document:
    df_print: paged
---

# Introduction

The statistical analysis has been carried in `r R.version$version.string`, version info, as well as used packages follow. If any of the needed packages are FALSE, please install them first.

## Packages

```{r echo=FALSE}
rm(list = ls())

package_list = c("yaml", "dplyr", "MASS", "car", "geepack", "ggplot2")

suppressPackageStartupMessages({
  sapply(package_list, require, character.only = TRUE)
})
```

## Configuration

Configuration is in the file `config.yml`.

```{r echo=FALSE}
config = yaml::read_yaml(file = "config.yml")
cat(paste0(readLines(con = "config.yml"), "\n"), sep = "")
```

```{r}
bc_data = read.csv(config$bc_data_f, header = TRUE, stringsAsFactors = FALSE)
```


# Data understanding

## Repeated measures

  1. Longitudinally bushcrickets have been measured twice (age 1 and age 2).
  2. For a given age, each bushcricket's response was measured to several different stimuli (six), however stimulus order has been randomized.
  3. For a fixed age, cricket, and stimulus, we have several repeated measures (up to 4).

*Repeated measures implies that there may be some correlation within the measurements of an individual.*

## Missing data

### Unbalanced between ages: different ages have different number of crickets in them.

Some bush-crickets died between ages, however their death was not correlated to any particular response, *therefore missing completely at random (MCAR)*. Several bush-crickets were recruited for age 2

```{r echo=FALSE}
cat("Number of measurements for age 1:")
bc_data %>% filter(age == 1) %>% nrow
cat("Number of measurements for age 2:")
bc_data %>% filter(age == 2) %>% nrow
```

### Unbalanced within age: also same bushcricket may be measured different times for a given age

Some stimuli were repeated less than 4 times because a bushcricket stopped singing. This does not seem to be random, however, as the bushcrickets that have some stimuli missing, tend to have also other missing.

In the printout below different columns represent different treatments but not necessarily in the same order. Shown informatively to inspect the balanceness of the data.

```{r echo=FALSE}
for (a in unique(bc_data$age)) {
  cat("Age", a, "\n")
 # print(unique(bc_data$stimuli))
  for (i in unique(bc_data$id)) {
    cat(i, " ")
    bc_data %>% dplyr::filter(age == a & id == i) %>% with({stimuli}) %>% table %>% cat
    cat("\n")
  }
}
rm(a, i)
```

With respect to the number of measurements, data does not appear to be missing at random. For this reason we have *summarized the up to four treatments,* see below.

## Transform variable stimuli into quantitaive form

| Type	| Frequency (kHz) |	dB SPL RMS @bushcricket	| Single call duration (ms)	| Pulse interval duration (ms)	| Total call train duration (ms) |
| ----------|-------|-----|---|-----|------|
| control   |	10	  | 88	| 3	| 30	| 1000 |
| bat call	| 85-25 |	87	| 3	| 30	| 1000 |
| bat call	| 85-25	| 77	| 3	| 30	| 1000 |
| bat call	| 85-25	| 67	| 3	| 30	| 1000 |
| bat call	| 85-25	| 57	| 3	| 30	| 1000 |
| silence   |	0	    |0	  | 0 |  0  | 1000 |

1. We create two variables `db_bat` (in the paper referred as "threat intensity") and `db_control`(in the paper referred as "control intensity") which contain the sound pressure level in decibels of the bat call and the control call.
2. We set `id` to be a factors.
3. Convert age to weeks

```{r}
do.call(
  function(bc_data,
           treatment_type = c("silence", "pt-very-low", "pt-low", "pt_moderate", "pt-high", "control"),
           db_bat = c(0, 57, 67, 77, 87, 0),
           db_control = c(0, 0, 0, 0, 0, 88),
           weeks = c(3,10))
  {
    bc_data$db_bat = db_bat[sapply(bc_data$stimuli, function(s) { which(treatment_type == s) })]
    bc_data$db_control = db_control[sapply(bc_data$stimuli, function(s) { which(treatment_type == s) })]
    bc_data$weeks = weeks[bc_data$age]
    bc_data$id = as.factor(bc_data$id)
    bc_data$age = as.factor(bc_data$age)
    return(bc_data)
  },
  list(bc_data)
) -> bc_data

bc_data$stimuli = factor(bc_data$stimuli, levels = c("silence", "pt-very-low", "pt-low", "pt_moderate", "pt-high", "control" ))
```

#### Average the treatments (id, stimuli, age) for each bush-cricket

- pause: round up the mean to nearest whole number in order to use a Poisson model
- reaction: is pause > 1 then yes, а single treatment is considered to have had reaction
- simultaneous_singing (in the paper referred as "early resume"): 
if the mean across repetitions is the less than 0 then yes

```{r}
bc_data %>% dplyr::group_by(id, stimuli, age) %>% dplyr::summarize(
                                                     reaction = (sum(pause > 1, na.rm =  TRUE) / length(pause)) >= .5,
                                                    pause = ceiling(mean(pause, na.rm =  TRUE)),
                                                    post.stimulus = ceiling(mean(post.stimulus, na.rm = TRUE)),
                                                    simultaneous_singing = ceiling(mean(post.stimulus, na.rm = TRUE)) < 0, 
                                                    db_bat = first(db_bat),
                                                    db_control = first(db_control),
                                                    weeks = first(weeks)
                                                  ) -> bc_data

```






# Modeling

## Marginal model (generalized estimating qquations)

### Basic

Idea:  model `pause` as a function of `age`, `db_bat` and `db_control`

```{r}
pause_f1 = pause ~ age + db_bat + db_control
pause_f2 = pause ~ age + db_bat + db_control + db_bat*age
pause_f3 = pause ~ age + stimuli                    
pause_f4 = pause ~ age + stimuli + age * stimuli    # with interaction

reaction_f1 =  reaction ~ age + db_bat + db_control
reaction_f2 =  reaction ~ age + db_bat + db_control + db_bat*age
reaction_f3 =  reaction ~ age + stimuli
reaction_f4 =  reaction ~ age + stimuli + age*stimuli
```

This supposes that the regression slope (but not necessarily the intercept) of
the marginal models will be the same. For example, for `age == 1` an increase in
10 dB in the bat call will lead to the same increase in response as for `age == 2`

- pause models the conditional parameter λ of the poisson distribution

Assess the conditional distributions Yi|Xi

- don't need to do it for reaction, as obviously Bernouili distributed

For pause see below

```{r, warning=FALSE, fig.keep='all'}
sapply(unique(bc_data$age), function(a) {
  sapply(unique(bc_data$db_bat), function(db) {
    sapply(unique(bc_data$db_control), function(dc) {
      pause = dplyr::filter(bc_data, age == a & db_bat == db & db_control == dc)$pause
      tryCatch({
        with(MASS::fitdistr(pause, "Poisson"), {
          car::qqp(pause, "pois", lambda = estimate)
          mtext(paste("age", a, "db_bat", db, "db_control", dc))
          
        })
        }, error = function(e) {
        #warning(e, " age", a, "db_bat ", db, "db_control ", dc, "\n")
      })
    })
  })
  return("")
})
```
# Model fit and selection

## GEE

- no need to specify random effects
- can deal with repeated measurements
- can deal with data, which is MCAR (missing completely at random)

```{r message=FALSE, warning=FALSE}
pause_m = list(
  geepack::geeglm(formula = pause_f1,
                   family = poisson("log"), 
                   corstr = "exchangeable", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = pause_f2,
                   family = poisson("log"), 
                   corstr = "exchangeable", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = pause_f3,
                   family = poisson("log"), 
                   corstr = "exchangeable", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = pause_f4,
                   family = poisson("log"), 
                   corstr = "exchangeable", data = bc_data,
                   id = id, waves = age),
 
  geepack::geeglm(formula = pause_f1,
                   family = poisson("log"), 
                   corstr = "independence", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = pause_f2,
                   family = poisson("log"), 
                   corstr = "independence", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = pause_f3,
                   family = poisson("log"), 
                   corstr = "independence", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = pause_f3,
                   family = poisson("log"), 
                   corstr = "independence", data = bc_data,
                   id = id, waves = age)
  
  )

reaction_m = list(
  geepack::geeglm(formula = reaction_f1,
                   family = binomial("logit"), 
                   corstr = "exchangeable", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = reaction_f2,
                   family = binomial("logit"), 
                   corstr = "exchangeable", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = reaction_f3,
                   family = binomial("logit"), 
                   corstr = "exchangeable", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = reaction_f4,
                   family = binomial("logit"), 
                   corstr = "exchangeable", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = reaction_f1,
                   family = binomial("logit"), 
                   corstr = "independence", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = reaction_f2,
                   family = binomial("logit"), 
                   corstr = "independence", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = reaction_f3,
                   family = binomial("logit"), 
                   corstr = "independence", data = bc_data,
                   id = id, waves = age),
  
  geepack::geeglm(formula = reaction_f4,
                   family = binomial("logit"), 
                   corstr = "independence", data = bc_data,
                   id = id, waves = age)
  
  )
```

## Model selection (GEE)

```{r}
options(digits=10)
sapply(pause_m, geepack::QIC)
sapply(reaction_m, geepack::QIC)
```

Preferred model: model 1 based on CIC
- no support for the hypothesis that there is an interaction
- prefer exchangebale correlation structure
- continuous covariates


# Results

## Model summaries 

- http://environmentalcomputing.net/interpreting-coefficients-in-glms/
- Model interpretation
- Controlling for multiple testing --- multivariate regression

```{r}
summary(pause_m[[1]])
p = tidy(pause_m[[1]])

summary(reaction_m[[1]])
r = tidy(reaction_m[[1]])

# corrected p.values
pvals = data.frame(p$p.value, r$p.value)#, s$p.value)
corrected_pvals = as.data.frame(do.call(rbind, lapply(1:nrow(pvals), function(i) {
  p.adjust(as.numeric(pvals[i, ]))
})))
names(corrected_pvals) = c("corr p", "corr r")#, "corr s")

p = cbind(p, corrected_pvals[,1])
r = cbind(r, corrected_pvals[,2])
p
r
```

## Assessing the model fit

pause_m[[1]] - preferred model, no interaction

```{r}
odds2p = function(odds) {
  exp(odds)/(1 + exp(odds))
}

dbs = c(87, 77, 67, 57, 0)
dbcs = c(0, 0, 0, 0, 88)
age = c(1, 2)

cat("Pause duration: observed vs modeled values")
sapply(
  1:length(dbs),
  function(ix)
  {
    sapply(age, function(a) 
    {
      m = mean(dplyr::filter(bc_data, age == a & db_bat == dbs[ix] & db_control == dbcs[ix])$pause)
      med = median(dplyr::filter(bc_data, age == a & db_bat == dbs[ix] & db_control == dbcs[ix])$pause)
      mod = exp(predict(pause_m[[1]], data.frame("age" = as.factor(a), db_bat = dbs[ix], db_control = dbcs[ix])))
      cat("age:", a, "bat:", dbs[ix], "control:", dbcs[ix], "mean:", m, "model:", mod, "\n")

    }
    )
    return ("")
  }
)

cat("Reaction probability: observed vs modeled values")
sapply(
  1:length(dbs),
  function(ix)
  {
    sapply(age, function(a) 
    {
      m = mean(dplyr::filter(bc_data, age == a & db_bat == dbs[ix] & db_control == dbcs[ix])$reaction)
      med = median(dplyr::filter(bc_data, age == a & db_bat == dbs[ix] & db_control == dbcs[ix])$reaction)
      mod =  odds2p(predict(pause_m[[1]], data.frame("age" = as.factor(a), db_bat = dbs[ix], db_control = dbcs[ix])))
      cat("age:", a, "bat:", dbs[ix], "control:", dbcs[ix], "mean:", m, "model:", mod, "\n")
    }
    )
    return("")
  }
)



```




## Calculate change of factor

```{r}
cat("Decrease of 10 db for age 1 leads to a change of factor for pause duration of:")
exp(predict(pause_m[[1]], data.frame("age" = as.factor(1), db_bat = 87, db_control = 0)))/
exp(predict(pause_m[[1]], data.frame("age" = as.factor(1), db_bat = 77, db_control = 0)))
```


## Graphs



```{r eval: FALSE}
bc_data$stimuli= factor(bc_data$stimuli, levels = c("pt-high", "pt_moderate", "pt-low", "pt-very-low", "silence", "control"))

bc_data$pause = bc_data$pause/1000
bc_data1 = bc_data
bc_data1[bc_data$pause == .001, "pause"] = .015

 bc_data   %>%
   dplyr::group_by(stimuli, age) %>% dplyr::summarize(reaction_percentage = sum(reaction)/length(reaction)) %>%
  ggplot( aes(x=stimuli, y = reaction_percentage, fill = age)) +
   geom_bar(stat = "identity", position = "dodge") + 
   scale_x_discrete(labels=c("pt-high" = "high", "pt_moderate" = "moderate",
                            "pt-low" = "low", "pt-very-low" = "very low", "control" = "cricket"))+
   xlab("Predation Threat") + ylab("Proportion of reacting males") -> p1a

 bc_data   %>%
   dplyr::group_by(stimuli, age) %>% dplyr::summarize(reaction_percentage = sum(reaction)/length(reaction)) %>%
  ggplot( aes(x=stimuli, y = reaction_percentage, fill = age, alpha = 0.5)) +
   geom_bar(stat = "identity", position = "dodge", width=0.8) +
 theme_bw()+
theme(text = element_text(size = 15), legend.position = "none",
      # remove the vertical grid lines
           panel.grid.major.x = element_blank() ,
           # explicitly set the horizontal lines (or they will disappear too)
           panel.grid.major.y = element_line( size=.01, color="grey" ) 
      ) +
  scale_fill_manual(values=c( "olivedrab1","olivedrab4"))+
   scale_x_discrete(labels=c("pt-high" = "High", "pt_moderate" = "Moderate",
                            "pt-low" = "Low", "pt-very-low" = "Very low", "silence" = "Silence",
                            "control" = "Control"))+
   xlab("Predation Threat") + ylab("Reacting males") -> p1
print(p1)
dev.print( png,filename="reacting-males.png",
           units="cm",width=14,res=300,bg="white")


bc_data  %>% dplyr::group_by(stimuli, age) %>% dplyr::summarize(ignore_threat = sum(simultaneous_singing)/length(simultaneous_singing)) %>%
 ggplot( aes(x=stimuli, y = ignore_threat, fill = age, alpha = 0.5)) +
  theme_bw()+
  theme(text = element_text(size = 15), 
        legend.position = "none",
        # remove the vertical grid lines
           panel.grid.major.x = element_blank() ,
           # explicitly set the horizontal lines (or they will disappear too)
           panel.grid.major.y = element_line( size=.01, color="grey" ) ) +
  scale_fill_manual(values=c( "olivedrab1","olivedrab4"))+
  geom_bar(stat = "identity", position = "dodge", width=0.8) +
  scale_x_discrete(labels=c("pt-high" = "High", "pt_moderate" = "Moderate",
                            "pt-low" = "Low", "pt-very-low" = "Very low", "silence" = "Silence",
                            "control" = "Control"))+
  xlab("Predation Threat") + ylab("Ignoring threat") -> p2
print(p2)
dev.print( png,filename="ignoring-predation-threat.png",
           units="cm",width=14,res=300,bg="white")


GeomSplitViolin <- ggproto("GeomSplitViolin", GeomViolin,
                           draw_group = function(self, data, ..., draw_quantiles = NULL) {
                             data <- transform(data, xminv = x - violinwidth * (x - xmin), 
                                               xmaxv = x + violinwidth * (xmax - x))
                             grp <- data[1, "group"]
                             newdata <- plyr::arrange(transform(data, x = if (grp %% 2 == 1) xminv else xmaxv), 
                                                      if (grp %% 2 == 1) y else -y)
                             newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
                             newdata[c(1, nrow(newdata) - 1, nrow(newdata)), "x"] <- round(newdata[1, "x"])

                             if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {
                               stopifnot(all(draw_quantiles >= 0), all(draw_quantiles <=
                                                                         1))
                               quantiles <- ggplot2:::create_quantile_segment_frame(data, draw_quantiles)
                               aesthetics <- data[rep(1, nrow(quantiles)), setdiff(names(data), c("x", "y")), 
                                                  drop = FALSE]
                               aesthetics$alpha <- rep(1, nrow(quantiles))
                               both <- cbind(quantiles, aesthetics)
                               quantile_grob <- GeomPath$draw_panel(both, ...)
                               ggplot2:::ggname("geom_split_violin", grid::grobTree(GeomPolygon$draw_panel(newdata, ...), quantile_grob))
                             }
                             else {
                               ggplot2:::ggname("geom_split_violin", GeomPolygon$draw_panel(newdata, ...))
                             }
                           })

geom_split_violin <- function(mapping = NULL, data = NULL, stat = "ydensity", position = "identity", ...,
                              draw_quantiles = NULL, trim = TRUE, scale = "area", na.rm = FALSE,
                              show.legend = NA, inherit.aes = TRUE) {
  layer(data = data, mapping = mapping, stat = stat, geom = GeomSplitViolin,
        position = position, show.legend = show.legend, inherit.aes = inherit.aes,
        params = list(trim = trim, scale = scale, draw_quantiles = draw_quantiles, na.rm = na.rm, ...))
}

bc_data1$pause[bc_data1$pause <.025] <- .025
f1 <- function(x) {
  log10(mean(10 ^ x))
}
 

ggplot(data = bc_data1, aes(x=stimuli, y = pause, fill = age)) +
  theme_classic()+
  scale_y_log10(breaks = c(.025,.050,.100, .200,.500,1.000,2.000, 5.000, 10.000, 20.000, 40.000), 
                limits = c(.024, 80)) +
  theme(panel.grid.minor = element_blank(),
         text = element_text(size = 16),
         legend.position = "none",
         # remove the vertical grid lines
           panel.grid.major.x = element_blank() ,
           # explicitly set the horizontal lines (or they will disappear too)
           panel.grid.major.y = element_line( size=.001, color="grey" )
    )+
  scale_fill_manual(values=c( "olivedrab1","olivedrab4"))+
  scale_color_manual(values=c( "olivedrab1","olivedrab4"))+
  scale_x_discrete( labels=c("pt-high" = "High", "pt_moderate" = "Moderate",
                           "pt-low" = "Low", "pt-very-low" = "Very low",
                           "silence" = "Silence", "control" = "Control"))+
  xlab("Predation threat") +
  ylab("Pause duration (s)")+
  geom_split_violin(trim = FALSE, alpha = 0.5, color = NA, adjust = .9, scale = "width")+
  geom_dotplot(binaxis = "y", stackdir = "center",
                stackratio=0.8, dotsize = 0.6,
                position = position_dodge(width = .5),
                alpha = 0.2,
               binwidth = 0.15) +
  stat_summary(fun= f1,
               geom = "crossbar",
               color=c("grey50","grey30","grey50","grey30","grey50","grey30",
                       "grey50","grey30","grey50","grey30","grey50","grey30"),
               width = 0.40,
               position = position_dodge(width = .40),) -> p4

print(p4)
dev.print( png,filename="pause_violin.png",
           units="cm",width=14,res=300,bg="white")

 


```

